## Repeated Measures Data

In his text _Regression Modeling Strategies_, Harrell writes (page 149 section 7.7):

> For an AR(1) correlation structure with $n$ equally spaced measurement times on $N$ subjects, with the correlation between two consecutive times being $\rho$, the effective sample size is $\frac{n-(n-2)\rho}{1+\rho}N$. Under compound symmetry, the effective sample size is $\frac{n}{1+\rho(n-1)}N$

```{r}
n <- 2 # number of repeated observations
r <- .8 # autocorrelation
N <- c(34,54) # number of participants (there are two groups)
# Two assumptions about correlation structure
ess.CS <- (n*N)/(1+r*(n-1))
ess.AR1 <- ((n-(n-2)*r)/(1+r))*N
# These will be the same because there are only 
# 2 repeated observations
ess.CS
ess.AR1
```

Now, to find the minimum detectable mean difference (with 5% type-I error level and
80% power using Lehr's rule) I will I will need to find the effective per-group
sample sizes, which is the harmonic mean of the two sample sizes

```{r}
hm.ess.CS <- 2/((1/ess.CS[1])+(1/ess.CS[2]))
hm.ess.CS
```

And now applying Lehr's rule

```{r}
d <- sqrt(16/hm.ess.CS)
d
```

One function to rule them all for two time point studies.

```{r}
minimum.d.ANCOVA <- function(n = 2, r,N) {
  # n <- number of repeated observations, default to 2
  # r <- autocorrelation
  # N <- number of participants (there are two groups), so N <- c(34,54)
  ess.CS <- (n*N)/(1+r*(n-1))
  hm.ess.CS <- 2/((1/ess.CS[1])+(1/ess.CS[2]))
  d <- sqrt(16/hm.ess.CS)
  d
}
minimum.d.ANCOVA(N=c(32,54),r=.8)
minimum.d.ANCOVA(N=c(85,54),r=.8)
```

